---
title: "Snippet 1"
author: "Ernesto Calvo"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

## Maximum Likelihood by Brute Force

This snippet provides an intuitive example of Maximum Likelihood by creating a matrix of loglik scores.

## Creating the data

We will first create a synthetic dataset that has known parameters:

```{r}
library("tidyverse")
library("plot3D")

# Set seed for reproducibility
set.seed(100)

# Define the sample size
T <- 1000

# Generate independent variable
x <- rnorm(T, 0, 1)

# We create a model that has parameters indexed by individual observation

#true_params_global <- list(alpha = 1.4, beta = 2.1, sigma = 1.0)
true_params <- list(alpha = rnorm(T,1.4), beta = rnorm(T,2.1), sigma = 1.0)


# Generate dependent variable
z <- rnorm(T, mean = true_params$alpha + true_params$beta * x, sd = true_params$sigma)

plot(z,x, cex=.4, pch=16, col="red")

```

Now we will retrieve the design parameters from the data.

First, let's try a brute force approach

```{r}

# Define grid for alpha and beta values
alpha_grid <- seq(-60, 60, by = 2) / 8
beta_grid <- seq(-160, 160, by = 4) / 20

# Compute log-likelihoods for each combination of alpha and beta values
ll_grid <- expand.grid(alpha = alpha_grid, beta = beta_grid) %>%
  mutate(log_likelihood = map2_dbl(alpha, beta, function(a, b) {
    sum(dnorm(z, mean = a + b * x, sd = 1, log = TRUE))
  })) %>%
  spread(beta, log_likelihood)

View(ll_grid)

```

The `echo: false` option disables the printing of code (only outp![Close document tab](data:image/png;base64,iVBORw0KGgoAAAANSUhEUgAAABIAAAASCAYAAABWzo5XAAAAWElEQVR42mNgGPTAxsZmJsVqQApgmGw1yApwKcQiT7phRBuCzzCSDSHGMKINIeDNmWQlA2IigKJwIssQkHdINgxfmBBtGDEBS3KCxBc7pMQgMYE5c/AXPwAwSX4lV3pTWwAAAABJRU5ErkJggg==){alt="Close document tab"}ut is displayed).

```{r}
# Find the optimal alpha and beta values
optimal_params <- ll_grid %>%
  gather(beta, log_likelihood, -alpha) %>%
  top_n(1, log_likelihood) %>%
  select(alpha, beta)

# Plot the log-likelihood surface
persp3D(alpha_grid, beta_grid, as.matrix(ll_grid[,-1]),
        theta = 3, phi = 20, axes = FALSE)

```

Explore the grid to find the set of parameters that maximize the likelihood of observing the data.

```{r}
max_location <- which(as.matrix(ll_grid[,-1]) == max(as.matrix(ll_grid[,-1])), arr.ind = TRUE)
print(max_location)
print(max(ll_grid))
#Estimate the model

# Define the negative log-likelihood function for a linear model
myreg <- function(theta, z, x) {
  sum(-dnorm(z,
             mean = theta[1] + theta[2] * x,
             sd = theta[3], log = TRUE))
}

# Estimate theta using the NLM function
estimate_theta <- function(z, x, theta_start = c(1, 1, 1), print_level = 2) {
  out <- nlm(myreg, theta_start, z = z, x = x, print.level = print_level, hessian = TRUE)
  
  tibble(
    MaximumLikelihood = -out$minimum,
    ParameterEstimates = out$estimate,
    StandardErrors = sqrt(diag(solve(out$hessian))),
    tvalues = out$estimate / sqrt(diag(solve(out$hessian))),
    Ratios = out$estimate / sqrt(diag(solve(out$hessian)))
  )
}

# Set initial values for the parameters
theta.start <- c(1, 1, 1)

# Estimate theta and display the results
results <- estimate_theta(z, x, theta_start = theta.start)
print(results)

```

```{r}
# Load libraries
library(ggplot2)

# Gather the results into key-value pairs
results_long <- results %>%
  select(ParameterEstimates, StandardErrors) %>%
  gather(key = "Variable", value = "Value")

# Create the ggplot
ggplot(results_long, aes(x = Variable, y = Value, color = Variable)) +
  geom_point(size = 3) +
  theme_minimal() +
  labs(title = "Parameter Estimates and Standard Errors",
       x = "Parameter",
       y = "Value")



# Calculate the confidence intervals
alpha <- 0.05 # 95% confidence interval
results_ci <- results %>%
  mutate(
    LowerBound = ParameterEstimates - qt(1 - alpha/2, T - 2) * StandardErrors,
    UpperBound = ParameterEstimates + qt(1 - alpha/2, T - 2) * StandardErrors
  ) %>%
  select(Parameter = ParameterEstimates, LowerBound, UpperBound) %>%
  gather(key = "Variable", value = "Value", -Parameter) %>%
  mutate(Variable = factor(Variable, levels = c("LowerBound", "Parameter", "UpperBound")))

plot(results$ParameterEstimates, ylim=c(0.5,2.5), ylab=c("alpha", "beta"))
segments(1:3,results_ci$Value,1:3, results_ci$Parameter)

```
